Simple connectome inference from partial correlation statistics in calcium imaging

### 2014-06-20 article.tex
small wording
blabla

The human brain is a complex biological organ made of 100 billions of neurons,
each connected to 7000 other neurons on average. Unfortunately, direct
observations of the connectome, the wiring diagram of the brain, is not yet
technically feasible. Without being perfect, calcium imaging currently allows
the real-time and simultaneous observation of neuron activities from thousands
of neurons, producing individual time series representing their fluorescence
intensity. From these data, the connectome inference problem consists in
retrieving the synaptic connections between neurons on the basis of the
fluorescence time series. In particular, this problem is often made difficult
because of experimental issues, including masking effects (making some of the
neurons not to be observed or confounded with others), the low sampling rate of
the optical device with respect to the neural activity speed or the slow decay
of fluorescence.

@@

Under the assumption that random variables follow a joint Gaussian
distribution $\mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and covariance
matrix $\Sigma$, $X_i$ and $X_j$ are conditionally independent with respect to
the other variables $X^{-i,j}$ if their \textit{partial correlation} statistic
is null~\citep{koller2009probabilistic}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j | X^{-i,j}} \neq 0$,
where $\rho_{X_i, X_j | X^{-i,j}}$ can now be interpreted as the degree of association
between the two variables (i.e., the larger the value, the stronger the interaction).
Together, the partial correlation statistics correspond to the inverse
$\Sigma^{-1}$ of the covariance matrix, also known as the concentration or
precision matrix. Computationally, they can be estimated efficiently in
$O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal components
extracted through PCA~\citep{bishop2006pattern} on the processed data. In our experiments, $M$ is set
to $0.8 p$, thereby allowing us to further filter noise in the data.

@@

Finally, let us remark that performance shows to be quite sensitive to the choice
of hyper-parameters in the signal processing pipeline (e.g., $f_1$ or $f_2$ or
the value of the threshold $\tau$). To further improve robustness, the last
step of our method therefore consists in averaging correlation statistics over
various values of the parameters, thereby reducing the variance of our
predictions. Further details are provided in Appendix~\ref{app:optimized}.

\textcolor{red}{G: Mention that correlation statistics are symmetric and do not account for direction.}

@@

\textcolor{red}{AJ: It's not clear on which dataset we focus on}

@@

\item \textbf{GENIE3} \citep{huynhthu2010inferring} translates the inference of the
network of $p$ neurons into $p$ supervised learning tasks. For each task,
the fluorescence of a neuron is predicted from the fluorescence
of all other neurons using a tree based ensemble such as
random forest \citep{breiman2001random}. The tree ensemble associate a
variable importance scores \citep{louppe2013understanding}
to each neuron used in the predictions, which is used as a connectivity
score. The following parameters are used
$10000$ trees, $\log_2{p}$ variables are randomly selected to split tree
nodes and the maximal tree depth is constrained to 1, 2 or 3.
\end{itemize}

@@

If the undirected network is perfectly inferred, it would already yield
high AUROC score for instance $\approx{}0.996$ on \textit{normal-1}. Trying
to infer edge directivity is likely to be counter-productive or yield
small gain.
By contrast, the same experiment with AUPRC would yield
to a score of $\approx 0.789$ on \textit{normal-1}. Thus inferring
correctly directivity would be more rewarding in AUPRC term.
Interestingly, methods that takes into account directivity, namely GTE and
GENIE3, improve their AUROC and AUPRC by giving the same way to directed edges.
(see GTE and GTE(undirected) in Figure \ref{fig:curves}
\textcolor{red}{AJ: Figure \ref{fig:curves}  contradicts this message,
especially pr-curve.}).
Trying to detect the link direction is more penalizing than rewarding.

@@

\begin{table}
\small
\centering
\begin{tabular}{@{}l *{8}{c}@{}}
  & \multicolumn{4}{c}{AUROC} & \multicolumn{4}{c}{AUPRC} \\
\hline \hline
Methods $\backslash$ normal- & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
\hline \hline
% Directivity \textcolor{red}{appendix?} & 0.535 & 0.526 & 0.533 & 0.535 & 0.012 & 0.012 & 0.012 & 0.012 \\
Correlation            & 0.886 & 0.884 & 0.891 & 0.877 & 0.153 & 0.145 & 0.170 & 0.132 \\
GENIE3 (depth = 1)     & 0.878 & & & & 0.184 & & & \\
GENIE3 (depth = 2)     & 0.889 & & & & 0.217 & & & \\
GENIE3 (depth = 3)     & & & & & & & & \\
GTE                    & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142 \\
Partial correlation    & 0.930 &  0.927 &  0.926 &  0.923& 0.363  & 0.347 &  0.346 & 0.328 \\
Partial correlation (avg.) & 0.938 & 0.936 & 0.936 & 0.932& 0.391 & 0.389 & 0.385 & 0.374\\
% Our approach simple + dir & 0.939 & 0.936 & 0.937 & 0.933& 0.392 & 0.390 & 0.386 & 0.375\\
Partial correlation (opt.) & 0.932 & 0.931 & 0.930 & 0.928 & 0.364 & 0.366 & 0.359 & 0.344 \\
Our approach (opt. \& avg.)    & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388 \\
% Our approach + dir  \textcolor{red}{appendix?}   & 0.944 & 0.943 & 0.942 & 0.940 & 0.404 & 0.405 & 0.399 & 0.389 \\
\end{tabular}
\caption{TODO}
\label{tab:tab1}
\end{table}

>>>>

\textcolor{red}{To improve (style):} The human brain is a complex biological organ made of 100 billions of neurons,
each connected to 7000 other neurons on average. Unfortunately, direct
observations of the connectome, the wiring diagram of the brain, is not yet
technically feasible. Without being perfect, calcium imaging currently allows
the real-time and simultaneous observation of neuron activities from thousands
of neurons, producing individual time series representing their fluorescence
intensity. From these data, the connectome inference problem consists in
retrieving the synaptic connections between neurons on the basis of the
fluorescence time series. In particular, this problem is often made difficult
because of experimental issues, including masking effects (making some of the
neurons not to be observed or confounded with others), the low sampling rate of
the optical device with respect to the neural activity speed or the slow decay
of fluorescence.

@@

Under the assumption that random variables follow a joint Gaussian
distribution $\mathcal{N}(\mu; \Sigma)$ of mean vector $\mu$ and covariance
matrix $\Sigma$, $X_i$ and $X_j$ are conditionally independent with respect to
the other variables $X^{-i,j}$ if their \textit{partial correlation} statistic
is null~\citep{koller2009probabilistic}, i.e., if
\begin{equation}
\rho_{X_i, X_j | X^{-i,j}} = -\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\end{equation}
Conversely, $X_i$ and $X_j$ are conditionally dependent if $\rho_{X_i, X_j | X^{-i,j}} \neq 0$,
where $\rho_{X_i, X_j | X^{-i,j}}$ can now be interpreted as the degree of association
between the two variables (i.e., the larger the value, the stronger the interaction).
Note however that the partial correlation statistic is symmetric (i.e., $\rho_{X_i, X_j | X^{-i,j}} = \rho_{X_j, X_i | X^{-j,i}}$) and therefore does not account for the direction of the interaction.
Together, the partial correlation statistics correspond to the inverse
$\Sigma^{-1}$ of the covariance matrix, also known as the concentration or
precision matrix. Computationally, they can be estimated efficiently in
$O(M^3)$ (instead of $O(p^3)$) from the $M$ first principal components
extracted through PCA~\citep{bishop2006pattern} on the processed data. In our experiments, $M$ is set
to $0.8 p$, thereby allowing us to further filter noise in the data.

@@

Finally, let us remark that performance shows to be quite sensitive to the choice
of hyper-parameters in the signal processing pipeline (e.g., $f_1$ or $f_2$ or
the value of the threshold $\tau$). To further improve robustness, the last
step of our method therefore consists in averaging correlation statistics over
various values of the parameters, thereby reducing the variance of our
predictions. In this case, the method is referred to as \textit{Averaged Partial correlation} statistics. Further details are provided in Appendix~\ref{app:optimized}.

@@

@@

\item \textbf{GENIE3} \citep{huynhthu2010inferring} translates the inference of the
network of $p$ neurons into $p$ supervised learning tasks. For each task,
the fluorescence of a neuron is predicted from the fluorescence
of all other neurons using a tree based ensemble such as
random forest \citep{breiman2001random}. The tree ensemble associates a
variable importance score \citep{louppe2013understanding}
to each neuron used in the predictions, which is used as a degree of associate. The following parameters are used
$10000$ trees, $\log_2{p}$ variables are randomly selected to split tree
nodes and the maximal tree depth is constrained to 1, 2 or 3.
\end{itemize}

@@

If the undirected network is perfectly inferred, it would already yield
high AUROC score for instance $\approx{}0.996$ on \textit{normal-1}. Trying
to infer edge directivity is likely to be counter-productive or yield
small gain.
By contrast, the same experiment with AUPRC would yield
to a score of $\approx 0.789$ on \textit{normal-1}. Thus inferring
correctly directivity would be more rewarding in AUPRC term.
Interestingly, methods that takes into account directivity, namely GTE and
GENIE3, improve their AUROC and AUPRC by giving the same way to directed edges.
(in particular see GTE and GTE(undirected) in Figure \ref{fig:pr_curve}
\textcolor{red}{AJ: Figure \ref{fig:curves}  contradicts this message,
especially pr-curve. AS:And now?} and Table \ref{tab:tab1} in Appendix \textcolor{red}{REF}).
Trying to detect the link direction is more penalizing than rewarding.

@@

\begin{table}
\small
\centering
\begin{tabular}{@{}l *{8}{c}@{}}
  & \multicolumn{4}{c}{AUROC} & \multicolumn{4}{c}{AUPRC} \\
\hline \hline
Methods $\backslash$ normal- & 1 & 2 & 3 & 4 & 1 & 2 & 3 & 4 \\
\hline \hline
% Directivity \textcolor{red}{appendix?} & 0.535 & 0.526 & 0.533 & 0.535 & 0.012 & 0.012 & 0.012 & 0.012 \\
Pearson correlation    & 0.886 & 0.884 & 0.891 & 0.877 & 0.153 & 0.145 & 0.170 & 0.132 \\
GENIE3 (depth = 1)     & 0.878 & & & & 0.184 & & & \\
GENIE3 (depth = 2)     & 0.889 & & & & 0.217 & & & \\
GENIE3 (depth = 3)     & & & & & & & & \\
GTE                    & 0.890 & 0.893 & 0.894 & 0.873 & 0.171 & 0.174 & 0.197 & 0.142 \\
Partial correlation    & 0.930 &  0.927 &  0.926 &  0.923& 0.363  & 0.347 &  0.346 & 0.328 \\
Partial correlation (avg.) & 0.938 & 0.936 & 0.936 & 0.932& 0.391 & 0.389 & 0.385 & 0.374\\
% Our approach simple + dir & 0.939 & 0.936 & 0.937 & 0.933& 0.392 & 0.390 & 0.386 & 0.375\\
Partial correlation (opt.) & 0.932 & 0.931 & 0.930 & 0.928 & 0.364 & 0.366 & 0.359 & 0.344 \\
Our approach (opt. \& avg.)    & 0.943 & 0.942 & 0.942 & 0.939 & 0.403 & 0.404 & 0.398 & 0.388 \\
% Our approach + dir  \textcolor{red}{appendix?}   & 0.944 & 0.943 & 0.942 & 0.940 & 0.404 & 0.405 & 0.399 & 0.389 \\
\end{tabular}
\caption{TODO}
\label{tab:tab1}
\end{table}
